---
title: "Midterm Project"
author: "Claire Sullivan"
format: html
editor: visual
---

## Loading Libraries

```{r}
library(tidyverse)
library(readr)
#install.packages("janitor")
library(janitor)
```

## Loading Data

```{r}
childcare_prices <- read_csv("childcare_prices_national.csv")
```

## Tidy Data

```{r}
# filter to California
childcare_prices_CA <- childcare_prices |> 
  filter(State_Name == "California") |> 
  mutate(County_Name = as.factor(County_Name))

# find columns where data is entered as character - sapply applies a function (in this case, is.character) over the vector of the column names for childcare_prices_CA

char_cols <- sapply(childcare_prices_CA, is.character)
# char_cols

# the only character columns we don't want to convert to numeric are State_Name, County_Name, State_Abbreviation, County_FIPS_Code

char_columns_to_convert <- 
  setdiff(names(childcare_prices_CA)[sapply(childcare_prices_CA, is.character)], c("State_Name", "County_Name", "State_Abbreviation", "County_FIPS_Code"))

childcare_prices_CA <- childcare_prices_CA |> 
  mutate(across(all_of(char_columns_to_convert), ~ as.numeric(gsub(",", "", gsub("\\$", "", .)))))

childcare_prices_CA
```

```{r}
# developing a metric for childcare burden by county - childcare burden will represent the median amount of money spent annually on childcare as a proportion of median household income for ONE parent with ONE child of any selected age group 

# also cleaning up data for employment sector columns because I want to look at that 

childcare_burden_CA <- childcare_prices_CA |> 
  group_by(StudyYear, County_Name) |> 
  drop_na(MCInfant, MCPreschool, MCToddler, MCSA) |> 
  mutate(
    annual_mc_infant = MCInfant * 52, # multiply by 52 to convert weekly to yearly
    annual_mc_preschool = MCPreschool * 52, 
    annual_mc_toddler = MCToddler * 52, 
    annual_mc_sa = MCSA * 52, 
    mc_infant_burden = annual_mc_infant/MHI, # yearly spending divided by yearly MHI
    mc_preschool_burden = annual_mc_preschool/MHI,
    mc_toddler_burden = annual_mc_toddler/MHI,
    mc_sa_burden = annual_mc_sa/MHI) |> 
  select(StudyYear, County_Name, contains("burden"), starts_with("EMP"), TotalPop) |> 
  rename(
    emp_mgmt_business_science_art = EMP_M, 
    emp_service = EMP_Service,
    emp_sales =  EMP_Sales, 
    emp_natlres_construct_maintenance = EMP_N,
    emp_production_transport_materialmoving = EMP_P
  ) |> 
  pivot_longer(-c(StudyYear, County_Name, starts_with("emp"), TotalPop), names_to = "age_group", values_to = "annual_childcare_burden") |> 
  pivot_longer(-c(StudyYear, County_Name, contains("burden"), age_group, annual_childcare_burden, TotalPop), names_to = "employment_sector", values_to = "percent_pop_employed_in_sector")

# make bins for population size
custom_breaks <- c(0, 100000, 500000, 1000000, Inf)
custom_labels <- c("1-99999", "100000-499999", "500000-1000000", "1000000+")

childcare_burden_CA <- childcare_burden_CA |> 
  mutate(population_bins = 
           cut(TotalPop, 
               breaks = custom_breaks, 
               labels = custom_labels))

```

```{r}
# dataset filtered down to just bay area counties 

# divide childcare cost annual/MHI ~ using within year divisions to avoid adjusting for inflation

bay_area_counties <- c("Alameda County", "Napa County", "Santa Clara County", "Contra Costa County", "San Francisco County", "Solano County", "Marin County", "San Mateo County", "Sonoma County")

bay_area_childcare <- childcare_prices_CA |> 
  filter(County_Name %in% bay_area_counties) |> 
  group_by(StudyYear, County_Name) |> 
  mutate(annual_mc_infant = MCInfant * 52,
         annual_mc_preschool = MCPreschool * 52, 
         annual_mc_toddler = MCToddler * 52, 
         annual_mc_sa = MCSA * 52, 
         mc_infant_burden = annual_mc_infant/MHI,
         mc_preschool_burden = annual_mc_preschool/MHI,
         mc_toddler_burden = annual_mc_toddler/MHI,
         mc_sa_burden = annual_mc_sa/MHI)


```

## Exploring Bay Area Counties

```{r}
bay_area_childcare2 <- bay_area_childcare |> 
  select(StudyYear, County_Name, mc_infant_burden, mc_preschool_burden, mc_toddler_burden, mc_sa_burden) |> 
  pivot_longer(-c(StudyYear, County_Name), names_to = "child_age", values_to = "burden")

bay_area_childcare2 |> 
  filter(County_Name %in% bay_area_counties, child_age == "mc_infant_burden") |> 
  group_by(County_Name) |> 
  ggplot(aes(x = StudyYear, y = burden, color = County_Name)) +
  geom_line() +
  labs(
    title = "Childcare Burden for Infant Center Based Care",
    x = "Year",
    y = "Childcare Burden",
    color = "County"
  )
```

```{r}
# comparing Median Household Income in Bay Area Counties
childcare_prices_CA |> 
  filter(County_Name %in% bay_area_counties, 
         StudyYear == 2018) |> 
  ggplot(aes(x = County_Name, y = MHI_2018, fill = County_Name)) +
  geom_col()

```

## Explore California Data

### Employment Sector

```{r}
childcare_burden_CA |> 
  filter(employment_sector == "emp_mgmt_business_science_art",
         StudyYear == "2018",
         age_group %in% c("mc_infant_burden", "mc_preschool_burden", "mc_sa_burden")) |> 
  ggplot(aes(x = percent_pop_employed_in_sector, y = annual_childcare_burden, color = age_group)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = .5) +
  labs(
    title = "Childcare Burden and Occupation in 2018",
    subtitle = "Employment in Management, Business, Science, and Art Occupations",
    x = "Percent of County Population Employed in Field",
    y = "Median Childcare Spending \nas Proportion of Household Income",
    color = "Age Group"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2", labels = c("Infant", "Toddler and Preschool", "School Age"))


childcare_burden_CA |> 
  filter(employment_sector == "emp_mgmt_business_science_art",
         StudyYear == "2018") 
```

```{r}
# statistical testing on employment sector 
childcare_burden_CA |>
  ungroup() |> 
  filter(StudyYear == "2018",
         employment_sector == "mc_mgmt_business_science_art")
```

```{r}
childcare_burden_CA |> 
  filter(age_group == "mc_preschool_burden") |> 
  ggplot(aes(x = percent_pop_employed_in_sector, y = annual_childcare_burden, group = employment_sector)) +
  geom_point(aes(color = employment_sector), alpha = 0.3)
```

```{r}
childcare_burden_CA |> 
  filter(age_group == "mc_preschool_burden") |> 
  ggplot(aes(x = percent_pop_employed_in_sector, y = annual_childcare_burden, color = employment_sector)) +
  geom_point()
```

```{r}
# trying another geom 
childcare_burden_CA |> 
  filter(StudyYear %in% c(2018), 
         age_group == "mc_preschool_burden") |> 
  ggplot(aes(x = percent_pop_employed_in_sector, y = annual_childcare_burden, fill = employment_sector)) +
  geom_violin(alpha = 0.3) +
  facet_grid(~StudyYear)
  
```

```{r}

# get quantiles 
childcare_burden_CA |> 
  group_by(employment_sector) |> 
  summarise(quantile = quantile(percent_pop_employed_in_sector))

employment_breaks <- c(21.8, 28.0, 32.7, 38.1, Inf)
employment_labels <- c("less than 21.8", "21.8-28%", "28-32.7%, 32.7-38.1%", "above 38%")

test2 <- childcare_burden_CA |> 
  filter(employment_sector == "emp_mgmt_business_science_art") |> 
  mutate(employment_bins = 
           cut(percent_pop_employed_in_sector,
               breaks = employment_breaks,
               labels = employment_labels)) |> 
  group_by(employment_bins) |> 
  summarise(avg_annual_burden = mean(annual_childcare_burden))

test2 |> 
  ggplot(aes(x = employment_bins, y = avg_annual_burden)) +
  geom_col()

```

### County Population Size Impacts

```{r}
# take weighted average of the medians by pouplation size in each county 

# weighted
childcare_burden_CA |> 
  group_by(population_bins) |> 
  summarise(weighted_mean = weighted.mean(x = annual_childcare_burden, w = TotalPop)) |> 
  ggplot(aes(x = population_bins, y = weighted_mean)) +
  geom_col()

# unweighted
childcare_burden_CA |> 
  group_by(population_bins) |> 
  summarise(average_burden = mean(annual_childcare_burden)) |> 
  ggplot(aes(x = population_bins, y = average_burden)) +
  geom_col()

```

```{r}
childcare_burden_CA |> 
  filter(age_group == "mc_infant_burden") |> 
  group_by(employment_sector)
```

```{r}
poverty_data_2018 <- read.csv("ACS_2018_poverty.csv")

poverty_data_2018 <- poverty_data_2018[-c(1), ] |> 
  separate(col = NAME, into = c("county", "state"), sep = ",")

poverty_data_2018 |> 
  mutate(state = as.character(state)) |> 
  filter(state == " California") |> 
  select()

```

## Creating Different Family Scenarios

-   two parents making median income with one kid

-   one parent making median income with one kid

-   two parents making median income with two kids

-   one parent making median income with two kids

```{r}
# using department of health and human services recommended 7% spending of income on childcare
two_parents_one_toddler <- childcare_burden_CA |> 
  group_by(StudyYear, County_Name) |> 
  filter(age_group == "mc_toddler_burden") |> 
  mutate(annual_toddler_burden_2_parents = annual_childcare_burden/(2)) |> 
  group_by(StudyYear, County_Name) |> 
  summarise(annual_burden = mean(annual_toddler_burden_2_parents)) |> 
  mutate(exceed_burden = annual_burden > 0.14)

two_parents_one_toddler |> 
 group_by(StudyYear) |>
  summarise(prop_exceed = mean(exceed_burden),
            prop_no_exceed = 1 - mean(exceed_burden)) |> 
  pivot_longer(-c(StudyYear), names_to = "condition", values_to = "proportion") |> 
  ggplot(aes(x = StudyYear, y = proportion, fill = condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()

test <- childcare_burden_CA |> 
  select(StudyYear, County_Name, population_bins, age_group, annual_childcare_burden) |> 
  group_by(StudyYear, County_Name) |> 
  mutate(exceed_burden = annual_childcare_burden > 0.07)
 
  
```
